Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
remind what an Experience Report is
explain Scored in the paper

\begin{abstract}
The Experience Report series is
a body of published, refereed, citable \emph{evidence} that functional
programming really works---or descriptions of obstacles that prevent
it from working.   
In~this Experience Report, we show that for computational biologists,
Haskell combines the flexibility and rapid prototyping of a scripting
language with performance comparable to~C++.
The Haskell language and libraries provide powerful abstractions, 
but we couldn't always get them to work!
Despite some speed bumps around profiling, debugging, and the
uncertain status of any given package, we~found Haskell's ecosystem
provided fine support for developing new algorithms in computational
biology.
\end{abstract}

@@

Our implementation computes not only the minimum
negated log probability, called a \texttt{Score},
but also the path of states which leads to that most likely outcome. 
Elsewhere, MRFy pairs not only paths but also other types of values
with scores, so we have defined a small abstraction:
\smallverbatiminput{vscore}
We~made \texttt{Scored} an instance of~\texttt{Ord} and of
\texttt{Functor}.

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state: 
we~choose the best solution
from among the three predecessor states; 
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
Programming the algorithm in this form was not only intellectually
satisfying; keeping the code close to the math also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

>>>>

%
% It can't hurt to remind everyone involved exactly what an Experience
% Report is.
%
%
\begin{abstract}
The Experience Report series is
a body of published, refereed, citable \emph{evidence} that functional
programming really works---or descriptions of obstacles that prevent
it from working.   
In~this Experience Report, we show that for computational biologists,
Haskell combines the flexibility and rapid prototyping of a scripting
language with performance comparable to~C++.
The Haskell language and libraries provide powerful abstractions, 
but we couldn't always get them to work!
Despite some speed bumps around profiling, debugging, and the
uncertain status of any given package, we~found Haskell's ecosystem
provided fine support for developing new algorithms in computational
biology.
\end{abstract}

@@

Our implementation computes not only the minimum
negated log probability, called a \texttt{Score},
but also the path of states which leads to that most likely outcome. 
Elsewhere, MRFy pairs not only paths but also other types of values
with scores, so we have defined a small abstraction:
\smallverbatiminput{vscore}
We~made \texttt{Scored} an instance of~\texttt{Ord} and of
\texttt{Functor}.
Think of a value of type \texttt{Scored~a} as a container holding
 an~``\texttt a'' with a number written on the side.
The \texttt{/+/} function adds to the number without touching the
 container.
Function 
\begin{smallverbatim}
fmap :: (a -> b) -> (Scored a -> Scored b)
\end{smallverbatim}
applies a function to what's inside the container, without touching
the number.
And containers are ordered purely by number, so applying
\texttt{minimum} to a list of scored things chooses the thing with the
smallest (and therefore best) score.

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
We~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
Programming the algorithm in this form was not only intellectually
satisfying; keeping the code close to the math also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.
